---
title: "Computational Musicology Portfolio"
author: "Matheus"
format: dashboard
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(compmus)

# ── Chord/key templates (Krumhansl-Kessler) ─────────────────────────────────

circshift <- function(v, n) {
  if (n == 0) v else c(tail(v, n), head(v, -n))
}

#      C     C#    D     Eb    E     F     F#    G     Ab    A     Bb    B
major_chord   <- c(1, 0, 0, 0, 1, 0, 0, 1, 0, 0, 0, 0)
minor_chord   <- c(1, 0, 0, 1, 0, 0, 0, 1, 0, 0, 0, 0)
seventh_chord <- c(1, 0, 0, 0, 1, 0, 0, 1, 0, 0, 1, 0)
major_key     <- c(6.35, 2.23, 3.48, 2.33, 4.38, 4.09, 2.52, 5.19, 2.39, 3.66, 2.29, 2.88)
minor_key     <- c(6.33, 2.68, 3.52, 5.38, 2.60, 3.53, 2.54, 4.75, 3.98, 2.69, 3.34, 3.17)

chord_templates <- tribble(
  ~name,       ~template,
  "Gb:7",      circshift(seventh_chord,  6),
  "Gb:maj",    circshift(major_chord,    6),
  "Bb:min",    circshift(minor_chord,   10),
  "Db:maj",    circshift(major_chord,    1),
  "F:min",     circshift(minor_chord,    5),
  "Ab:7",      circshift(seventh_chord,  8),
  "Ab:maj",    circshift(major_chord,    8),
  "C:min",     circshift(minor_chord,    0),
  "Eb:7",      circshift(seventh_chord,  3),
  "Eb:maj",    circshift(major_chord,    3),
  "G:min",     circshift(minor_chord,    7),
  "Bb:7",      circshift(seventh_chord, 10),
  "Bb:maj",    circshift(major_chord,   10),
  "D:min",     circshift(minor_chord,    2),
  "F:7",       circshift(seventh_chord,  5),
  "F:maj",     circshift(major_chord,    5),
  "A:min",     circshift(minor_chord,    9),
  "C:7",       circshift(seventh_chord,  0),
  "C:maj",     circshift(major_chord,    0),
  "E:min",     circshift(minor_chord,    4),
  "G:7",       circshift(seventh_chord,  7),
  "G:maj",     circshift(major_chord,    7),
  "B:min",     circshift(minor_chord,   11),
  "D:7",       circshift(seventh_chord,  2),
  "D:maj",     circshift(major_chord,    2),
  "F#:min",    circshift(minor_chord,    6),
  "A:7",       circshift(seventh_chord,  9),
  "A:maj",     circshift(major_chord,    9),
  "C#:min",    circshift(minor_chord,    1),
  "E:7",       circshift(seventh_chord,  4),
  "E:maj",     circshift(major_chord,    4),
  "G#:min",    circshift(minor_chord,    8),
  "B:7",       circshift(seventh_chord, 11),
  "B:maj",     circshift(major_chord,   11),
  "D#:min",    circshift(minor_chord,    3)
)

key_templates <- tribble(
  ~name,       ~template,
  "Gb:maj",    circshift(major_key,   6),
  "Bb:min",    circshift(minor_key,  10),
  "Db:maj",    circshift(major_key,   1),
  "F:min",     circshift(minor_key,   5),
  "Ab:maj",    circshift(major_key,   8),
  "C:min",     circshift(minor_key,   0),
  "Eb:maj",    circshift(major_key,   3),
  "G:min",     circshift(minor_key,   7),
  "Bb:maj",    circshift(major_key,  10),
  "D:min",     circshift(minor_key,   2),
  "F:maj",     circshift(major_key,   5),
  "A:min",     circshift(minor_key,   9),
  "C:maj",     circshift(major_key,   0),
  "E:min",     circshift(minor_key,   4),
  "G:maj",     circshift(major_key,   7),
  "B:min",     circshift(minor_key,  11),
  "D:maj",     circshift(major_key,   2),
  "F#:min",    circshift(minor_key,   6),
  "A:maj",     circshift(major_key,   9),
  "C#:min",    circshift(minor_key,   1),
  "E:maj",     circshift(major_key,   4),
  "G#:min",    circshift(minor_key,   8),
  "B:maj",     circshift(major_key,  11),
  "D#:min",    circshift(minor_key,   3)
)

# ── Load data ────────────────────────────────────────────────────────────────
diw_chroma  <- read_csv("compmus_audio/do_i_wanna_know.csv",  show_col_types = FALSE)
chop_chroma <- read_csv("compmus_audio/chop_suey.csv",        show_col_types = FALSE)
diw_timbre  <- read_csv("compmus_audio/do_i_wanna_know_timbre.csv", show_col_types = FALSE)
chop_timbre <- read_csv("compmus_audio/chop_suey_timbre.csv",       show_col_types = FALSE)
```

# Corpus

## Row {height=25%}

### Most-liked track {.value-box icon="music-note-beamed"}

**Do I Wanna Know?** — Arctic Monkeys
285 plays · ~7% skip rate

### Most-skipped track {.value-box icon="skip-forward-fill"}

**Chop Suey!** — System of a Down
100 plays · 100% skip rate

## Row {height=75%}

### About this corpus {width=55%}

This portfolio analyses two tracks that sit at opposite ends of my Spotify listening behaviour:
**Do I Wanna Know?** by Arctic Monkeys, which I have played 285 times and almost never skip,
and **Chop Suey!** by System of a Down, which I consistently skip immediately.

The contrast is deliberate: one track is hypnotic and repetitive (you can put it on and zone out),
the other is confrontational and unpredictable (it demands your full attention or you bail).
Computational features should reflect this difference clearly at every level of analysis —
from raw pitch content to timbre to harmonic structure.

### Track details {width=45%}

| Feature | Do I Wanna Know? | Chop Suey! |
|---|---|---|
| Artist | Arctic Monkeys | System of a Down |
| Genre | Indie rock | Nu-metal |
| Duration | ~4:26 | ~3:29 |
| Key | E minor / G major | D minor (drop-D) |
| Tempo | ~85 bpm | Variable |
| Play count | 285 | ~100 |
| Skip rate | ~7% | 100% |

# Chromagrams

## Row

### Do I Wanna Know? {width=50%}

```{r}
#| label: chromagram-diw
diw_chroma |>
  compmus_wrangle_chroma() |>
  mutate(pitches = map(pitches, compmus_normalise, "manhattan")) |>
  compmus_gather_chroma() |>
  mutate(pitch_class = factor(
    pitch_class, levels = c("C","C#","D","Eb","E","F","F#","G","Ab","A","Bb","B")
  )) |>
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", name = "Magnitude") +
  scale_x_continuous(name = "Time (s)", expand = expansion(mult = 0)) +
  labs(title = "Do I Wanna Know? — Chromagram", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold"))
```

### Chop Suey! {width=50%}

```{r}
#| label: chromagram-chop
chop_chroma |>
  compmus_wrangle_chroma() |>
  mutate(pitches = map(pitches, compmus_normalise, "manhattan")) |>
  compmus_gather_chroma() |>
  mutate(pitch_class = factor(
    pitch_class, levels = c("C","C#","D","Eb","E","F","F#","G","Ab","A","Bb","B")
  )) |>
  ggplot(aes(x = start + duration / 2, width = duration, y = pitch_class, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", name = "Magnitude") +
  scale_x_continuous(name = "Time (s)", expand = expansion(mult = 0)) +
  labs(title = "Chop Suey! — Chromagram", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold"))
```

## Row {height=20%}

### Interpretation

**Do I Wanna Know?** shows a very stable chromagram: E, G, and A dominate from start to finish,
matching the song's signature riff which barely moves harmonically.
**Chop Suey!** is far more energetic — chroma energy jumps between pitch classes and there
are clear shifts at ~60 s, ~120 s, and ~155 s, corresponding to the song's abrupt dynamic drops
and explosions. The spread of energy across pitch classes also reflects drop-D tuning and a wider harmonic vocabulary.

# Cepstrograms

## Row

### Do I Wanna Know? {width=50%}

```{r}
#| label: cepstrogram-diw
diw_timbre |>
  compmus_wrangle_timbre() |>
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", name = "Magnitude") +
  scale_x_continuous(name = "Time (s)", expand = expansion(mult = 0)) +
  labs(title = "Do I Wanna Know? — Cepstrogram", y = "MFCC coefficient") +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold"))
```

### Chop Suey! {width=50%}

```{r}
#| label: cepstrogram-chop
chop_timbre |>
  compmus_wrangle_timbre() |>
  mutate(timbre = map(timbre, compmus_normalise, "manhattan")) |>
  compmus_gather_timbre() |>
  ggplot(aes(x = start + duration / 2, width = duration, y = mfcc, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", name = "Magnitude") +
  scale_x_continuous(name = "Time (s)", expand = expansion(mult = 0)) +
  labs(title = "Chop Suey! — Cepstrogram", y = "MFCC coefficient") +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(), plot.title = element_text(face = "bold"))
```

## Row {height=20%}

### Interpretation

**Do I Wanna Know?** is timbrally stable: low MFCCs (mfcc_02 to mfcc_04) hold steady throughout,
reflecting a consistent guitar tone and production. There is a drop around 130 s where the song
strips back to guitar only. **Chop Suey!** is all over the place — mfcc_02 and mfcc_03 spike
hard during heavy sections and collapse during the quiet bridge (~100–140 s).
That bridge is acoustically a completely different world from the rest of the song,
and the cepstrogram captures the contrast precisely.

# Structure

## Row

### Do I Wanna Know? — Chroma SSM {width=25%}

```{r}
#| label: ssm-diw-chroma
diw_chroma |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "manhattan")) |>
  compmus_self_similarity(pitches, "manhattan") |>
  filter(!is.na(d)) |>
  mutate(d = d / max(d)) |>
  ggplot(aes(
    x = xstart + xduration / 2, width = 50 * xduration,
    y = ystart + yduration / 2, height = 50 * yduration,
    fill = d
  )) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(title = "Do I Wanna Know? — Chroma SSM", x = "Time (s)", y = "Time (s)") +
  theme(plot.title = element_text(face = "bold"))
```

### Do I Wanna Know? — Timbre SSM {width=25%}

```{r}
#| label: ssm-diw-timbre
diw_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  filter(!is.na(d)) |>
  mutate(d = d / max(d)) |>
  ggplot(aes(
    x = xstart + xduration / 2, width = 50 * xduration,
    y = ystart + yduration / 2, height = 50 * yduration,
    fill = d
  )) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(title = "Do I Wanna Know? — Timbre SSM", x = "Time (s)", y = "Time (s)") +
  theme(plot.title = element_text(face = "bold"))
```

### Chop Suey! — Chroma SSM {width=25%}

```{r}
#| label: ssm-chop-chroma
chop_chroma |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(pitches = map(pitches, compmus_normalise, "manhattan")) |>
  compmus_self_similarity(pitches, "manhattan") |>
  filter(!is.na(d)) |>
  mutate(d = d / max(d)) |>
  ggplot(aes(
    x = xstart + xduration / 2, width = 50 * xduration,
    y = ystart + yduration / 2, height = 50 * yduration,
    fill = d
  )) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(title = "Chop Suey! — Chroma SSM", x = "Time (s)", y = "Time (s)") +
  theme(plot.title = element_text(face = "bold"))
```

### Chop Suey! — Timbre SSM {width=25%}

```{r}
#| label: ssm-chop-timbre
chop_timbre |>
  compmus_wrangle_timbre() |>
  filter(row_number() %% 50L == 0L) |>
  mutate(timbre = map(timbre, compmus_normalise, "euclidean")) |>
  compmus_self_similarity(timbre, "cosine") |>
  filter(!is.na(d)) |>
  mutate(d = d / max(d)) |>
  ggplot(aes(
    x = xstart + xduration / 2, width = 50 * xduration,
    y = ystart + yduration / 2, height = 50 * yduration,
    fill = d
  )) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(guide = "none") +
  theme_classic() +
  labs(title = "Chop Suey! — Timbre SSM", x = "Time (s)", y = "Time (s)") +
  theme(plot.title = element_text(face = "bold"))
```

## Row {height=20%}

### Interpretation

**Do I Wanna Know?** shows clear repetitive block structure in both SSMs: diagonal blocks mark
internally consistent verses and choruses, and bright off-diagonal rectangles confirm the same
sections returning with the same instrumentation. **Chop Suey!** is much more fragmented —
small, uneven blocks reflect the song's many short distinct sections.
The quiet bridge (~100–140 s) appears as a dark isolated island in the timbre SSM because
nothing else in the track sounds remotely like it.

# Harmony

## Row

### Do I Wanna Know? — Keygram {width=50%}

```{r}
#| label: keygram-diw
diw_chroma |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, "euclidean", "manhattan") |>
  ggplot(aes(
    x = start + duration / 2, width = 50 * duration,
    y = name, fill = d
  )) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  theme_minimal() +
  labs(
    title = "Do I Wanna Know? — Keygram",
    subtitle = "Krumhansl-Kessler profiles · euclidean/manhattan",
    x = "Time (s)", y = NULL
  ) +
  theme(
    plot.title    = element_text(face = "bold"),
    panel.grid    = element_blank()
  )
```

### Chop Suey! — Keygram {width=50%}

```{r}
#| label: keygram-chop
chop_chroma |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(key_templates, "euclidean", "manhattan") |>
  ggplot(aes(
    x = start + duration / 2, width = 50 * duration,
    y = name, fill = d
  )) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  theme_minimal() +
  labs(
    title = "Chop Suey! — Keygram",
    subtitle = "Krumhansl-Kessler profiles · euclidean/manhattan",
    x = "Time (s)", y = NULL
  ) +
  theme(
    plot.title    = element_text(face = "bold"),
    panel.grid    = element_blank()
  )
```

## Row

### Do I Wanna Know? — Chordogram {width=50%}

```{r}
#| label: chordogram-diw
diw_chroma |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(chord_templates, "euclidean", "manhattan") |>
  ggplot(aes(
    x = start + duration / 2, width = 50 * duration,
    y = name, fill = d
  )) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  theme_minimal() +
  labs(
    title = "Do I Wanna Know? — Chordogram",
    subtitle = "1-0 chord templates · euclidean/manhattan",
    x = "Time (s)", y = NULL
  ) +
  theme(
    plot.title    = element_text(face = "bold"),
    panel.grid    = element_blank()
  )
```

### Chop Suey! — Chordogram {width=50%}

```{r}
#| label: chordogram-chop
chop_chroma |>
  compmus_wrangle_chroma() |>
  filter(row_number() %% 50L == 0L) |>
  compmus_match_pitch_template(chord_templates, "euclidean", "manhattan") |>
  ggplot(aes(
    x = start + duration / 2, width = 50 * duration,
    y = name, fill = d
  )) +
  geom_tile() +
  scale_fill_viridis_c(guide = "none") +
  theme_minimal() +
  labs(
    title = "Chop Suey! — Chordogram",
    subtitle = "1-0 chord templates · euclidean/manhattan",
    x = "Time (s)", y = NULL
  ) +
  theme(
    plot.title    = element_text(face = "bold"),
    panel.grid    = element_blank()
  )
```

## Row {height=25%}

### Key findings

**Do I Wanna Know? — key stability.** The keygram is dominated by a narrow band in the
E minor / G major region throughout the entire song. This is exactly what you would expect:
the track is built around a single repeating riff that cycles through Em, G, Am, and D,
never really modulating. The chordogram reinforces this — E:min, G:maj, and A:min stand out
persistently, corresponding to the three main chords of the riff. The harmonic vocabulary is
deliberately minimal; the song's appeal comes from groove and timbre, not harmonic variety.

**Chop Suey! — harmonic ambiguity.** The keygram is far messier. D minor is the most
consistent presence (consistent with drop-D tuning and the song's centre of gravity), but
competing regions light up during the bridge and outro. The chordogram shows rapid,
erratic switching between chord areas, which matches the song's aggressive rhythmic and
harmonic interruptions. Neither the key nor the chord estimation settles cleanly — the template
matching algorithm is genuinely confused, and that confusion is musically accurate:
Chop Suey! deliberately avoids the kind of tonal stability that makes pattern matching easy.
